Study guide: Software engineering with exponential decay models

### 2015-10-10 doc/src/chapters/softeng/slides_softeng.tex

% LaTeX Beamer file automatically generated from DocOnce
% https://github.com/hplgit/doconce

%-------------------- begin beamer-specific preamble ----------------------

\documentclass{beamer}

\usetheme{red_shadow}
\usecolortheme{default}

% turn off the almost invisible, yet disturbing, navigation symbols:
\setbeamertemplate{navigation symbols}{}

% Examples on customization:
%\usecolortheme[named=RawSienna]{structure}
%\usetheme[height=7mm]{Rochester}
%\setbeamerfont{frametitle}{family=\rmfamily,shape=\itshape}
%\setbeamertemplate{items}[ball]
%\setbeamertemplate{blocks}[rounded][shadow=true]
%\useoutertheme{infolines}
%
%\usefonttheme{}
%\useinntertheme{}
%
%\setbeameroption{show notes}
%\setbeameroption{show notes on second screen=right}

% fine for B/W printing:
%\usecolortheme{seahorse}

\usepackage{pgf}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{relsize}

\usepackage{fancybox}  % make sure fancybox is loaded before fancyvrb

\usepackage{fancyvrb}
\usepackage{minted} % requires pygments and latex -shell-escape filename
%\usepackage{anslistings}
%\usepackage{listingsutf8}

\usepackage{amsmath,amssymb,bm}
%\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{colortbl}
\usepackage[english]{babel}
\usepackage{tikz}
\usepackage{framed}
% Use some nice templates
\beamertemplatetransparentcovereddynamic

% --- begin table of contents based on sections ---
% Delete this, if you do not want the table of contents to pop up at
% the beginning of each section:
% (Only section headings can enter the table of contents in Beamer
% slides generated from DocOnce source, while subsections are used
% for the title in ordinary slides.)
\AtBeginSection[]
{
  \begin{frame}<beamer>[plain]
  \frametitle{}
  %\frametitle{Outline}
  \tableofcontents[currentsection]
  \end{frame}
}
% --- end table of contents based on sections ---

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\newcommand{\shortinlinecomment}[3]{\note{\textbf{#1}: #2}}
\newcommand{\longinlinecomment}[3]{\shortinlinecomment{#1}{#2}{#3}}

\definecolor{linkcolor}{rgb}{0,0,0.4}
\hypersetup{
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3
    }
\setlength{\parskip}{7pt}  % {1em}

\newenvironment{doconceexercise}{}{}
\newcounter{doconceexercisecounter}
\newenvironment{doconce:movie}{}{}
\newcounter{doconce:movie:counter}

\newcommand{\subex}[1]{\noindent\textbf{#1}}  % for subexercises: a), b), etc

%-------------------- end beamer-specific preamble ----------------------

% Add user's preamble

@@

\title{Study guide: Software engineering with exponential decay models}

@@

\author{Hans Petter Langtangen\inst{1,2}}
\institute{Center for Biomedical Computing, Simula Research Laboratory\inst{1}
\and
Department of Informatics, University of Oslo\inst{2}}
% ----------------- end author(s) -------------------------

\date{Oct 10, 2015
% <optional titlepage figure>
% <optional copyright>
}

\begin{frame}[plain,fragile]
\titlepage
\end{frame}

\section{From flat program to module with functions}

\begin{frame}[plain,fragile]
\frametitle{Mathematical model problem}

@@

$\theta =0$: Forward Euler, $\theta =1$: Backward Euler, $\theta =1/2$: Crank-Nicolson (midpoint method)
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Many will make a rough, flat program first}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
from numpy import *
from matplotlib.pyplot import *

@@

y_e = A*exp(-a*t) - y
error = y_e - y
E = sqrt(dt*sum(error**2))
print 'Norm of the error: %.3E' % E
plot(t, y, 'r--o')
t_e = linspace(0, T, 1001)
y_e = A*exp(-a*t_e)
plot(t_e, y_e, 'b-')
legend(['numerical, theta=%g' % theta, 'exact'])
xlabel('t')
ylabel('y')
show()
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{There are major issues with this solution}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{New flat program}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
from numpy import *
from matplotlib.pyplot import *

@@

# Compare numerical (u) and exact solution (u_e) in a plot
plot(t, u, 'r--o')
t_e = linspace(0, T, 1001)       # very fine mesh for u_e
u_e = I*exp(-a*t_e)
plot(t_e, u_e, 'b-')
legend(['numerical, theta=%g' % theta, 'exact'])
xlabel('t')
ylabel('u')
show()
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Such flat programs are ideal for IPython notebooks!}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{fig-softeng/ipynb_flat.png}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{But: Further development of such flat programs require many scattered edits - easy to make mistakes!}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
def solver(I, a, T, dt, theta):
    """Solve u'=-a*u, u(0)=I, for t in (0,T] with steps of dt."""
    dt = float(dt)               # avoid integer division
    Nt = int(round(T/dt))        # no of time intervals
    T = Nt*dt                    # adjust T to fit time step dt
    u = np.zeros(Nt+1)           # array of u[n] values
    t = np.linspace(0, T, Nt+1)  # time mesh

    u[0] = I                  # assign initial condition
    for n in range(0, Nt):    # n=0,1,...,Nt-1
        u[n+1] = (1 - (1-theta)*a*dt)/(1 + theta*dt*a)*u[n]
    return u, t
\end{minted}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
u, t = solver(I=1, a=2, T=4, dt=0.2, theta=0.5)
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The DRY principle: Don't repeat yourself!}

\begin{block}{DRY: }
When implementing a particular functionality in a computer program, make sure
this functionality and its variations are implemented in just one piece
of code. That is, if you need to revise the implementation, there should be
\emph{one and only one} place to edit. It follows that you should never
duplicate code (don't repeat yourself!), and code snippets that are
similar should be factored into one piece (function) and parameterized (by
function arguments).
\end{block}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Make sure any program file is a valid Python module}

@@

\noindent
\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
from decay import solver
# Solve a decay problem
u, t = solver(I=1, a=2, T=4, dt=0.2, theta=0.5)
\end{minted}
or prefix function names by the module name:

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
import decay
# Solve a decay problem
u, t = decay.solver(I=1, a=2, T=4, dt=0.2, theta=0.5)
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The requirements of a module are so simple}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
if __name__ == '__main__':
    # Statements
\end{minted}

If the file is imported, the if test fails and no main program is run,
otherwise, the file works as a program
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The module file \texttt{decay.py} for our example}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
from numpy import *
from matplotlib.pyplot import *

@@

if __name__ == '__main__':
    experiment_compare_numerical_and_exact()
\end{minted}

Complete file: \href{{http://tinyurl.com/ofkw6kc/softeng/decay.py}}{\nolinkurl{decay.py}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{The module file \texttt{decay.py} for our example w/prefix}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
import numpy as np
import matplotlib.pyplot as plt

@@

if __name__ == '__main__':
    experiment_compare_numerical_and_exact()
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{How do we add code for comparing schemes visually?}

@@

Think of edits in the flat program that are required to produce this plot (!)
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{We just add a new function with the tailored plotting}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
def experiment_compare_schemes():
    """Compare theta=0,1,0.5 in the same plot."""
    I = 1;  a = 2;  T = 4;  dt = 0.4
    legends = []
    for theta in [0, 1, 0.5]:
        u, t = solver(I, a, T, dt, theta)
        plt.plot(t, u, '--o')
        legends.append('theta=%g' % theta)
    t_e = np.linspace(0, T, 1001)        # very fine mesh for u_e
    u_e = u_exact(t_e, I, a)
    plt.plot(t_e, u_e, 'b-')
    legends.append('exact')
    plt.legend(legends, loc='upper right')
    plotfile = 'tmp'
    plt.savefig(plotfile + '.png');  plt.savefig(plotfile + '.pdf')

@@

        logging.info('u[%d]=%g' % (n, u[n]))
        logging.debug('1 - (1-theta)*a*dt: %g, %s' %
                      (1-(1-theta)*a*dt,
                       str(type(1-(1-theta)*a*dt))[7:-2]))
        logging.debug('1 + theta*dt*a: %g, %s' %
                      (1 + theta*dt*a,
                       str(type(1 + theta*dt*a))[7:-2]))
    return u, t
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Prefixing imported functions by the module name}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
from numpy import *
from matplotlib.pyplot import *
\end{minted}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
import numpy as np
import matplotlib.pyplot as plt
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Documenting functions and modules}

\label{softeng1:basic:docstring}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Example on NumPy-style doc string}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
def solver(I, a, T, dt, theta):
    """
    Solve :math:`u'=-au` with :math:`u(0)=I` for :math:`t \in (0,T]`
    with steps of `dt` and the method implied by `theta`.

@@

    >>> u, t = solver(I=1.5, a=0.5, T=9, theta=0.5)
    >>> import matplotlib.pyplot as plt
    >>> plt.plot(t, u)
    >>> plt.show()
    """
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Example on an autogenerated nice HTML manual}

@@

\noindent
% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-softeng/selfdoc_numpy.png}}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Logging intermediate results}

\label{softeng1:basic:logging}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Introductory example on using a logger}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
import logging
import logging
logging.basicConfig(
    filename='myprog.log', filemode='w', level=logging.WARNING,
    format='%(asctime)s - %(levelname)s - %(message)s',
    datefmt='%m/%d/%Y %I:%M:%S %p')
logging.info('Here is some general info.')
logging.warning('Here is a warning.')
logging.debug('Here is some debugging info.')
logging.critical('Dividing by zero!')
logging.error('Encountered an error.')
\end{minted}

@@

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{text}
09/26/2015 09:25:10 AM - INFO - Here is some general info.
09/26/2015 09:25:10 AM - WARNING - Here is a warning.
09/26/2015 09:25:10 AM - CRITICAL - Dividing by zero!
09/26/2015 09:25:10 AM - ERROR - Encountered an error.
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{A message is tied to a level, and one can specify one many levels that get printed}

@@

\noindent
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Using a logger in our solver function}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=2mm]{python}
import logging
logging.basicConfig(
    filename='decay.log', filemode='w', level=logging.DEBUG,
    format='%(asctime)s - %(levelname)s - %(message)s',
    datefmt='%Y.%m.%d %I:%M:%S %p')

@@

        logging.info('u[%d]=%g' % (n, u[n]))
        logging.debug('1 - (1-theta)*a*dt: %g, %s' %
                      (1-(1-theta)*a*dt,
                       str(type(1-(1-theta)*a*dt))[7:-2]))
        logging.debug('1 + theta*dt*a: %g, %s' %
                      (1 + theta*dt*a,
                       str(type(1 + theta*dt*a))[7:-2]))
    return u, t
\end{minted}
\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Monitoring messages}

@@

\end{frame}

\begin{frame}[plain,fragile]
\frametitle{Accessing command-line arguments}

@@

\noindent
\end{frame}

>>>>

%%
%% Automatically generated file from DocOnce source
%% (https://github.com/hplgit/doconce/)
%%

% #define PREAMBLE

% #ifdef PREAMBLE
%-------------------- begin preamble ----------------------

\documentclass[%
oneside,                 % oneside: electronic viewing, twoside: printing
final,                   % draft: marks overfull hboxes, figures with paths
10pt]{article}

\listfiles               % print all files needed to compile this document

\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
\usepackage[table]{xcolor}
\usepackage{bm,microtype}

\usepackage[pdftex]{graphicx}

\usepackage{fancybox}  % make sure fancybox is loaded before fancyvrb
%\setlength{\fboxsep}{8pt}  % may clash with need in pre/cod envirs

% Packages for typesetting blocks of computer code
\usepackage{fancyvrb,framed,moreverb}

% Define colors
\definecolor{orange}{cmyk}{0,0.4,0.8,0.2}
\definecolor{tucorange}{rgb}{1.0,0.64,0}
\definecolor{darkorange}{rgb}{.71,0.21,0.01}
\definecolor{darkgreen}{rgb}{.12,.54,.11}
\definecolor{myteal}{rgb}{.26, .44, .56}
\definecolor{gray}{gray}{0.45}
\definecolor{mediumgray}{gray}{.8}
\definecolor{lightgray}{gray}{.95}
\definecolor{brown}{rgb}{0.54,0.27,0.07}
\definecolor{purple}{rgb}{0.5,0.0,0.5}
\definecolor{darkgray}{gray}{0.25}
\definecolor{darkblue}{rgb}{0,0.08,0.45}
\definecolor{darkblue2}{rgb}{0,0,0.8}
\definecolor{lightred}{rgb}{1.0,0.39,0.28}
\definecolor{lightgreen}{rgb}{0.48,0.99,0.0}
\definecolor{lightblue}{rgb}{0.53,0.81,0.92}
\definecolor{lightblue2}{rgb}{0.3,0.3,1.0}
\definecolor{lightpurple}{rgb}{0.87,0.63,0.87}
\definecolor{lightcyan}{rgb}{0.5,1.0,0.83}

\colorlet{comment_green}{green!50!black}
\colorlet{string_red}{red!60!black}
\colorlet{keyword_pink}{magenta!70!black}
\colorlet{indendifier_green}{green!70!white}

% Backgrounds for code
\definecolor{cbg_gray}{rgb}{.95, .95, .95}
\definecolor{bar_gray}{rgb}{.92, .92, .92}

\definecolor{cbg_yellowgray}{rgb}{.95, .95, .85}
\definecolor{bar_yellowgray}{rgb}{.95, .95, .65}

\colorlet{cbg_yellow2}{yellow!10}
\colorlet{bar_yellow2}{yellow!20}

\definecolor{cbg_yellow1}{rgb}{.98, .98, 0.8}
\definecolor{bar_yellow1}{rgb}{.98, .98, 0.4}

\definecolor{cbg_red1}{rgb}{1, 0.85, 0.85}
\definecolor{bar_red1}{rgb}{1, 0.75, 0.85}

\definecolor{cbg_blue1}{rgb}{0.87843, 0.95686, 1.0}
\definecolor{bar_blue1}{rgb}{0.7,     0.95686, 1}
%\setlength{\fboxsep}{-1.5mm}  % do not change since !bbox needs it positive!

%% Background for code blocks (parameter is color name)

%% pro/cod_vpad: gives some vertical padding before and after the text
%% (but has more simplistic code than _cod/pro_tight+cod/pro).
%% pro/cod_vpad can be used to enclose Verbatim or lst begin/end for code.
%% pro/cod calls _pro/cod_tight and has very little vertical padding,
%% used to enclose Verbatim and other begin/end for code.
%% (pro/cod is what the ptex2tex program could produce with the
%% Blue/BlueBar definitions in .ptex2tex.cfg.)

\newenvironment{cod_vpad}[1]{
   \def\FrameCommand{\colorbox{#1}}
   \MakeFramed{\FrameRestore}}
   {\endMakeFramed}

\newenvironment{_cod_tight}[1]{
   \def\FrameCommand{\colorbox{#1}}
   \FrameRule0.6pt\MakeFramed {\FrameRestore}\vskip3mm}
   {\vskip0mm\endMakeFramed}

\newenvironment{cod}[1]{
\bgroup\rmfamily
\fboxsep=0mm\relax
\begin{_cod_tight}{#1}
\list{}{\parsep=-2mm\parskip=0mm\topsep=0pt\leftmargin=2mm
\rightmargin=2\leftmargin\leftmargin=4pt\relax}
\item\relax}
{\endlist\end{_cod_tight}\egroup}

%% Background for complete program blocks (parameter 1 is color name
%% for background, parameter 2 is color for left bar)
\newenvironment{pro_vpad}[2]{
   \def\FrameCommand{\color{#2}\vrule width 1mm\normalcolor\colorbox{#1}}
   \MakeFramed{\FrameRestore}}
   {\endMakeFramed}

\newenvironment{_pro_tight}[2]{
   \def\FrameCommand{\color{#2}\vrule width 1mm\normalcolor\colorbox{#1}}
   \FrameRule0.6pt\MakeFramed {\advance\hsize-2mm\FrameRestore}\vskip3mm}
   {\vskip0mm\endMakeFramed}

\newenvironment{pro}[2]{
\bgroup\rmfamily
\fboxsep=0mm\relax
\begin{_pro_tight}{#1}{#2}
\list{}{\parsep=-2mm\parskip=0mm\topsep=0pt\leftmargin=2mm
\rightmargin=2\leftmargin\leftmargin=4pt\relax}
\item\relax}
{\endlist\end{_pro_tight}\egroup}

\usepackage{listingsutf8}

% Common lstlisting parameters
\lstset{
  basicstyle=\small \ttfamily,
  breaklines=false,          % break/wrap lines
  breakatwhitespace=true,    % let linebreaks happen at whitespace
  breakindent=40pt,
  tab=,
  tabsize=4,                 % tab means 4 spaces
  %belowskip=\smallskipamount,  % space between code and text below
  xleftmargin=5pt,           % indentation of code frame
  xrightmargin=5pt,
  framexleftmargin=5pt,      % add frame space to the left of code
  %numbers=left,             % put line numbers on the left
  %stepnumber=2,             % stepnumber=1 numbers each line, =n every n lines
  %framerule=0.4pt           % thickness of frame
  aboveskip=2ex,             % vertical space above code frame
  showstringspaces=false,    % show spaces in strings with an underscore
  showspaces=false,          % show spaces with an underscore
  showtabs=false,
  keepspaces=true,
  columns=fullflexible,      % tighter character kerning, like verb
  escapeinside={||},         % for |\pause| in slides and math in code blocks
  extendedchars=\true,       % allows non-ascii chars, does not work with utf-8
}

% Internally defined styles for lstlisting

\lstdefinestyle{simple}{
commentstyle={},
}

% end of custom lstdefinestyles

\usepackage[T1]{fontenc}
%\usepackage[latin1]{inputenc}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}

\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern

% Hyperlinks in PDF:
\definecolor{linkcolor}{rgb}{0,0,0.4}
\usepackage{hyperref}
\hypersetup{
    breaklinks=true,
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    citecolor=black,
    filecolor=black,
    %filecolor=blue,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
    }
%\hyperbaseurl{}   % hyperlinks are relative to this root

\setcounter{tocdepth}{2}  % number chapter, section, subsection

% Tricks for having figures close to where they are defined:
% 1. define less restrictive rules for where to put figures
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{0.75}
% floatpagefraction must always be less than topfraction!
% 2. ensure all figures are flushed before next section
\usepackage[section]{placeins}
% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
%\usepackage{float}\restylefloat{figure}

\usepackage[framemethod=TikZ]{mdframed}

% --- begin definitions of admonition environments ---

% --- end of definitions of admonition environments ---

% prevent orhpans and widows
\clubpenalty = 10000
\widowpenalty = 10000

\newenvironment{doconceexercise}{}{}
\newcounter{doconceexercisecounter}

% ------ header in subexercises ------
%\newcommand{\subex}[1]{\paragraph{#1}}
%\newcommand{\subex}[1]{\par\vspace{1.7mm}\noindent{\bf #1}\ \ }
\makeatletter
% 1.5ex is the spacing above the header, 0.5em the spacing after subex title
\newcommand\subex{\@startsection*{paragraph}{4}{\z@}%
                  {1.5ex\@plus1ex \@minus.2ex}%
                  {-0.5em}%
                  {\normalfont\normalsize\bfseries}}
\makeatother

% --- end of standard preamble for documents ---

@@

\thispagestyle{empty}

\begin{center}
{\LARGE\bf
\begin{spacing}{1.25}
Study guide: Software engineering with exponential decay models
\end{spacing}
}
\end{center}

@@

\begin{center}
{\bf Hans Petter Langtangen${}^{1, 2}$} \\ [0mm]
\end{center}

\begin{center}
% List of all institutions:
\centerline{{\small ${}^1$Center for Biomedical Computing, Simula Research Laboratory}}
\centerline{{\small ${}^2$Department of Informatics, University of Oslo}}
\end{center}

% ----------------- end author(s) -------------------------

% --- begin date ---
\begin{center}
Oct 10, 2015
\end{center}
% --- end date ---

\vspace{1cm}

% !split
\section*{From flat program to module with functions}

% !split
\subsection*{Mathematical model problem}

@@

$\theta =0$: Forward Euler, $\theta =1$: Backward Euler, $\theta =1/2$: Crank-Nicolson (midpoint method)

% !split
\subsection*{Many will make a rough, flat program first}

\begin{pro}{cbg_yellow2}{bar_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
from numpy import *
from matplotlib.pyplot import *

@@

y_e = A*exp(-a*t) - y
error = y_e - y
E = sqrt(dt*sum(error**2))
print 'Norm of the error: %.3E' % E
plot(t, y, 'r--o')
t_e = linspace(0, T, 1001)
y_e = A*exp(-a*t_e)
plot(t_e, y_e, 'b-')
legend(['numerical, theta=%g' % theta, 'exact'])
xlabel('t')
ylabel('y')
show()
\end{lstlisting}\end{pro}
\noindent

% !split

\subsection*{There are major issues with this solution}

@@

\noindent
% !split
\subsection*{New flat program}

\begin{pro}{cbg_yellow2}{bar_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
from numpy import *
from matplotlib.pyplot import *

@@

# Compare numerical (u) and exact solution (u_e) in a plot
plot(t, u, 'r--o')
t_e = linspace(0, T, 1001)       # very fine mesh for u_e
u_e = I*exp(-a*t_e)
plot(t_e, u_e, 'b-')
legend(['numerical, theta=%g' % theta, 'exact'])
xlabel('t')
ylabel('u')
show()
\end{lstlisting}\end{pro}
\noindent

% !split
\subsection*{Such flat programs are ideal for IPython notebooks!}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{fig-softeng/ipynb_flat.png}}

% !split
\subsection*{But: Further development of such flat programs require many scattered edits - easy to make mistakes!}

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
def solver(I, a, T, dt, theta):
    """Solve u'=-a*u, u(0)=I, for t in (0,T] with steps of dt."""
    dt = float(dt)               # avoid integer division
    Nt = int(round(T/dt))        # no of time intervals
    T = Nt*dt                    # adjust T to fit time step dt
    u = np.zeros(Nt+1)           # array of u[n] values
    t = np.linspace(0, T, Nt+1)  # time mesh

    u[0] = I                  # assign initial condition
    for n in range(0, Nt):    # n=0,1,...,Nt-1
        u[n+1] = (1 - (1-theta)*a*dt)/(1 + theta*dt*a)*u[n]
    return u, t
\end{lstlisting}\end{cod}
\noindent

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
u, t = solver(I=1, a=2, T=4, dt=0.2, theta=0.5)
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{The DRY principle: Don't repeat yourself!}

% --- begin paragraph admon ---
\paragraph{DRY:}
When implementing a particular functionality in a computer program, make sure
this functionality and its variations are implemented in just one piece
of code. That is, if you need to revise the implementation, there should be
\emph{one and only one} place to edit. It follows that you should never
duplicate code (don't repeat yourself!), and code snippets that are
similar should be factored into one piece (function) and parameterized (by
function arguments).
% --- end paragraph admon ---

% !split
\subsection*{Make sure any program file is a valid Python module}

@@

\noindent
\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
from decay import solver
# Solve a decay problem
u, t = solver(I=1, a=2, T=4, dt=0.2, theta=0.5)
\end{lstlisting}\end{cod}
\noindent
or prefix function names by the module name:

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
import decay
# Solve a decay problem
u, t = decay.solver(I=1, a=2, T=4, dt=0.2, theta=0.5)
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{The requirements of a module are so simple}

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
if __name__ == '__main__':
    # Statements
\end{lstlisting}\end{cod}
\noindent

If the file is imported, the if test fails and no main program is run,
otherwise, the file works as a program

% !split
\subsection*{The module file \texttt{decay.py} for our example}

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
from numpy import *
from matplotlib.pyplot import *

@@

if __name__ == '__main__':
    experiment_compare_numerical_and_exact()
\end{lstlisting}\end{cod}
\noindent

Complete file: \href{{http://tinyurl.com/ofkw6kc/softeng/decay.py}}{\nolinkurl{decay.py}}

% !split
\subsection*{The module file \texttt{decay.py} for our example w/prefix}

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
import numpy as np
import matplotlib.pyplot as plt

@@

if __name__ == '__main__':
    experiment_compare_numerical_and_exact()
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{How do we add code for comparing schemes visually?}

@@

Think of edits in the flat program that are required to produce this plot (!)

% !split
\subsection*{We just add a new function with the tailored plotting}

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
def experiment_compare_schemes():
    """Compare theta=0,1,0.5 in the same plot."""
    I = 1;  a = 2;  T = 4;  dt = 0.4
    legends = []
    for theta in [0, 1, 0.5]:
        u, t = solver(I, a, T, dt, theta)
        plt.plot(t, u, '--o')
        legends.append('theta=%g' % theta)
    t_e = np.linspace(0, T, 1001)        # very fine mesh for u_e
    u_e = u_exact(t_e, I, a)
    plt.plot(t_e, u_e, 'b-')
    legends.append('exact')
    plt.legend(legends, loc='upper right')
    plotfile = 'tmp'
    plt.savefig(plotfile + '.png');  plt.savefig(plotfile + '.pdf')

@@

        logging.info('u[%d]=%g' % (n, u[n]))
        logging.debug('1 - (1-theta)*a*dt: %g, %s' %
                      (1-(1-theta)*a*dt,
                       str(type(1-(1-theta)*a*dt))[7:-2]))
        logging.debug('1 + theta*dt*a: %g, %s' %
                      (1 + theta*dt*a,
                       str(type(1 + theta*dt*a))[7:-2]))
    return u, t
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{The comparison}

% !split
\subsection*{Prefixing imported functions by the module name}

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
from numpy import *
from matplotlib.pyplot import *
\end{lstlisting}\end{cod}
\noindent

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
import numpy as np
import matplotlib.pyplot as plt
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{Documenting functions and modules}
\label{softeng1:basic:docstring}

@@

\noindent
% !split
\subsection*{Example on NumPy-style doc string}

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
def solver(I, a, T, dt, theta):
    """
    Solve :math:`u'=-au` with :math:`u(0)=I` for :math:`t \in (0,T]`
    with steps of `dt` and the method implied by `theta`.

@@

    >>> u, t = solver(I=1.5, a=0.5, T=9, theta=0.5)
    >>> import matplotlib.pyplot as plt
    >>> plt.plot(t, u)
    >>> plt.show()
    """
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{Example on an autogenerated nice HTML manual}

@@

\noindent
% inline figure
\centerline{\includegraphics[width=0.8\linewidth]{fig-softeng/selfdoc_numpy.png}}

% !split
\subsection*{Logging intermediate results}
\label{softeng1:basic:logging}

@@

\noindent
% !split
\subsection*{Introductory example on using a logger}

\begin{pro}{cbg_yellow2}{bar_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
import logging
import logging
logging.basicConfig(
    filename='myprog.log', filemode='w', level=logging.WARNING,
    format='%(asctime)s - %(levelname)s - %(message)s',
    datefmt='%m/%d/%Y %I:%M:%S %p')
logging.info('Here is some general info.')
logging.warning('Here is a warning.')
logging.debug('Here is some debugging info.')
logging.critical('Dividing by zero!')
logging.error('Encountered an error.')
\end{lstlisting}\end{pro}
\noindent

@@

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
09/26/2015 09:25:10 AM - INFO - Here is some general info.
09/26/2015 09:25:10 AM - WARNING - Here is a warning.
09/26/2015 09:25:10 AM - CRITICAL - Dividing by zero!
09/26/2015 09:25:10 AM - ERROR - Encountered an error.
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{A message is tied to a level, and one can specify one many levels that get printed}

@@

\noindent
% !split
\subsection*{Using a logger in our solver function}

\begin{cod}{cbg_yellow2}\begin{lstlisting}[language=Python,style=simple,xleftmargin=2mm]
import logging
logging.basicConfig(
    filename='decay.log', filemode='w', level=logging.DEBUG,
    format='%(asctime)s - %(levelname)s - %(message)s',
    datefmt='%Y.%m.%d %I:%M:%S %p')

@@

        logging.info('u[%d]=%g' % (n, u[n]))
        logging.debug('1 - (1-theta)*a*dt: %g, %s' %
                      (1-(1-theta)*a*dt,
                       str(type(1-(1-theta)*a*dt))[7:-2]))
        logging.debug('1 + theta*dt*a: %g, %s' %
                      (1 + theta*dt*a,
                       str(type(1 + theta*dt*a))[7:-2]))
    return u, t
\end{lstlisting}\end{cod}
\noindent

% !split
\subsection*{Monitoring messages}

@@

\noindent
% !split
\subsection*{Accessing command-line arguments}

@@

\noindent
% !split
\subsection*{Working with an argument parser}
